SPENCER THESIS - WORK IN PROGRESS

### 2016-05-30 doc/thesis_swe.tex
Thesis update.

% While the effects of gravitational lensing can be dramatic, the shape of most galaxies is only distorted by a few percent and must be detected statistically as the intrinsic shape is not known.

Predicting accurate dark matter mass maps from weak lensing requires understanding the correlation between galaxies and dark matter structure. Numerous N-body simulations have predicted dark matter structure to have evolved from overdense regions in the early universe into `halos' of virialized matter that eventually collapsed from self-gravity (cite), where galaxies could then form in the large gravitational potentials of the halos. However, there is still extensive reserach being done on the correct density profile of the halos (isothermal spheres \cite{isothermal_spheres}, NFW \cite{nfw}, BMO \cite{bmo}, Einasto \cite{einasto}), identifying the stellar mass-halo mass relation \cite{smhr}, abundance matching \cite{reddick}, halo substructure \cite{substructure}, and the lensing effects of line of sight halo structure \cite{collet} and dark matter structures not associated with galaxies \cite{mccully}. If a model for the distribution of dark matter in a region of foreground mass can accurately predict the statistical signal of the weak lensing of background sources, then the model can be used on galaxies in existing sky survey data to extrapolate the amount of dark matter in the region and construct large-scale maps of the dark matter in the universe.

Guided by these results, in this thesis I build upon the work of Collet et al. in \cite{collet} to apply a simple dark matter halo model to reconstruct lensing mass along lines of sight in the Millennium Simulation using the \texttt{Pangloss} framework in order to predict the weakly lensed ellipticities of generated background sources. I then compare my predicted lensed ellipticities to the `true' lensing calculated by Hilbert et al. in \cite{ray_tracing} by ray-tracing light through the Millennium Simulation. By doing so, I attempt to answer the following questions: Under the best possible circumstances of having noiseless redshifts and halo masses, how accurately can I reconstruct line-of-sight dark matter mass structures using a halo model? What fraction of dark matter mass is in halos? How feasible is an inference of the Millennium Simulation's cosmological hyper-parameters, such as the stellar mass-halo mass relation, using all available halo data? How long does such an inference take?

A brief introduction to the halo model, the effects of strong and weak gravitational lensing, representing galaxy ellipticities, and cosmological correlation functions is discussed in Section \ref{background}. The implementation of the \texttt{Pangloss} framework to make line of sight lensing predictions in the Millennium Simulation is described in Section \ref{pangloss}, and the results of the framework on ?? galaxies on a field size of ?? arcmin$^2$, along with a comparison of the predicted lensed ellipticities to ray-traced ellipticities, is given in Section \ref{check}. Section \ref{feasibility} explores the feasibility of using \texttt{Pangloss} to make parameter inferences using all available halo data either by likelihood estimation or approximate Bayesian computation. Section \ref{discussion} discusses limitations of the used framework as well as potential physical and computational improvements that can be made for upcoming work before concluding remarks in Section \ref{conclusion}.

@@

Cosmological simulations like the Millennium Simulation (MS) demonstrate that the underlying dark matter distribution of our universe is likely breathtakingly complex; rather than a largely disordered collection of galaxies and interstellar gas, matter appears to condense into clusters and superclusters of galaxies threaded together by large filaments (pocketed) with under-dense regions called voids, as shown in Figure \ref{fig:ms_structure} (Cite MS). While the exact relation between the distribution of galaxies and dark matter is not known, simulations predict that early galaxies formed in the large gravitational potentials of over-dense regions of dark matter (Citation). This suggests that, while we cannot map dark matter directly, galaxies should trace out at least some of the underlying dark matter structures (Citation). A comparrison of the predicted galaxy and dark matter distribution from a small region of the MS is shown in Figure \ref{ms_compare}.

@@

The simplest way to model the relationship between galaxies and the dark matter distribution is by enveloping each galaxy in a spherically symmetric dark matter `halo' of a certain mass $M_h$. These halos extend far beyond the edge of the visible galaxy that they enclose, with the Milky Way's own halo estimated to have a radius of (Check and cite). While the density profile of the halos may be complex, numerous simulations have shown that it can be well approximated locally by the Navarro-Frenk-White (NFW) profile which has the form
\begin{equation}\label{nfw_profile}
\rho_{NFW}(r)=\frac{\rho_0}{\frac{r}{R_s}\left(1+\frac{r}{R_s}\right)^2}
\end{equation}

\noindent where the constant $\rho_0$ and the scale radius $R_s$ are parameters that vary from halo to halo \cite{nfw}. However, the NFW profile is not entierly physical as the total mass diverges as the desntiy is integrated out to an infinite radius, so this work uses a truncated NFW profile called the Baltz-Marshall-Oguri (BMO) profile given by
\begin{equation}\label{bmo_profile}
\rho_{BMO}(r)=\rho_{NFW}\cdot\left(\frac{r_t^2}{r^2+r_t^2}\right)^2
\end{equation}

@@

\noindent where $d_H=c/H_0$ is the Hubble distance, $\Omega_k$ is the curvature parameter ($\Omega_k=0$ for a flat universe), and $d_C(z)$ is the line-of-sight (LOS) comoving distance given by
\begin{equation}\label{comoving_distance}
d_C(z)=d_H\int_0^z\frac{dz^\prime}{E(z^\prime)}
\end{equation}

@@

As the LOS proper distance between two objects $r_p(z)$ is related to the LOS comoving distance $d_C(z)$ by
\begin{equation}\label{physical2comoving}
r_p(z)=a(z)d_C(z)=\frac{d_H}{1+z}\int_0^z\frac{dz^\prime}{E(z^\prime)},
\end{equation}

@@

\noindent Combining this result with Equation \eqref{temp_solution} leads to the desired proper volume element given by
\begin{equation}\label{pvolume_element}
dV_p=\frac{d_Hd_M(z)^2}{E(z)(1+z)^3}dzd\Omega.
\end{equation}

@@

A full mathematical treatment of the gravitational lensing of galaxies due to the gravitational fields of massive objects requires general relativity (see \cite{modern_cosmology} for details). However, the important results can be summarized as follows. Foreground mass distorts the image of a background galaxy in two distinct ways: The image is magnified and sheared tangentially about the foreground mass, making it more elliptical. The magnification of the image is determined by the convergence $\kappa$ defined by
\begin{equation}\label{convergence}
\kappa(\vec{\theta})=\frac{\sum(\vec{r},z_l)}{\sum_{cr}(z_l,z_s)}
\end{equation}

@@

However, usually the quantity of interest in lensing calculations is the reduced shear, defined as
\begin{equation}\label{reduced_shear}
g=\frac{\gamma}{1-\kappa}.
\end{equation}

@@

 The analytical form of the convergence and shear contribution from a BMO-profile is quite messy, but is derived and explained in detail in (Cite Phil's paper!!). Calculating the lensing contribution of the smooth-component correction is much simpler; the projected surface mass density at at given redshift is found by integrating $\rho_s(z)$ over a sufficiently small $dz$ interval, and then the mass sheet's convergence is found using Equation \eqref{kappa}. There is no shear contribution as a uniform sheet of mass causes only magnification due to the mass sheet degeneracy (CITE!). However, Equation \eqref{lensing} shows that the lensed ellipticities depend on the \textit{reduced} shear, which does depend on the additional convergence terms resulting from the smooth-component correction.

@@

To calculate the lensed ellipticities in the previous section, galaxies must first be translated from pixel intensities into elliptical representations of the galaxies. This is not always possible for galaxies are diverse in type and shape and not all are well approximated by ellipses. However for this thesis, all galaxy ellipticities are simulated so this issue is avoided. For a more thorough analysis of measuring image ellipticities, see Schneider (2003) \cite{schneider}.

@@

\noindent and $r\leq1$ is the ratio of the semi-minor and semi-major axes of the ellipse. This compact notation combines the eccentricity and orientation of the ellipse into a single object. A plot from Schneider (2003) \cite{schneider} showing the shape of elliptical galaxies for various values of $\varepsilon_1$ and $\varepsilon_2$ is shown in Figure \ref{ellipses}.

@@

If all galaxies were spherical, then the measurement of their ellipticities would give an exact description of the lensing done by foreground mass structures. Unfortunately, the intrinsic ellipticities of background galaxies contribute significant noise as we do not know how any individual galaxy ellipticity has been altered by lensing. However as two sources lensed by the same foreground mass will have their ellipticities distorted (in a similar way), their lensed ellipticities should be correlated regardless of intrinsic orientation. The average correlation of a sufficient number of galaxy pairs should overcome the noise and lead to a detectable signal. Importantly, this implies that \textit{the correlation of galaxy ellipticities is a probe of the underlying dark matter mass distribution} (Citation!).

@@

As galaxies will only be lensed by the same foreground mass at relatively small separation distances, the $\varepsilon$-$\varepsilon$ correlation function decreases sharply as the separation increases. If the distribution of intrinsic galaxy ellipticities is indeed random, then their $\varepsilon$-$\varepsilon$ correlation function should be essentially zero across all separation scales.

A similar statistical measure often used in ... is the galaxy-mass correlation function, which measures ... and is defined as
\begin{equation}\label{ng_corr_def}
\xi=
\end{equation}

\noindent where...

@@

To infer foreground dark matter mass structures using weak gravitational lensing, first a model of the relationship between foreground galaxies and the foreground dark matter must be established and robustly tested to see if, statistically, it makes the same lensing predictions of background sources as the true underlying dark matter structure. To do this, I built upon the publicly available \texttt{Pangloss}\footnote{\label{note1}https://github.com/drphilmarshall/Pangloss} framework used in Collett et al. \cite{collett_marshall} to reconstruct all the
mass along the lines of sight of generated background galaxies in the MS using a halo model with smooth-component correction. The lensing contribution of each halo is calculated, and the total lensing of the background galaxy is the sum of each halo contribution. The implementation of this process is detailed in the following sections.

@@

Testing the first assumption is the main goal of this paper. The second assumption is reasonable as it is estimated that dark matter halos are on average 1-2 orders of magnitude more massive than the host galaxy \cite{smhr}. Clearly the third assumption will never be true for any observational data. However, this allows for a best-case estimate of how well the \texttt{Pangloss} framework \textit{could} predict weak lensing effects given all possible information. This assumption can be relaxed by sampling a dark matter halo mass from an assumed stellar mass - halo mass relation. The fourth assumption is also unrealistic as most galaxies in sky surveys only have a less-reliable photometric redshift due to time constraints, but this again allows for a best-case estimate. This assumption could be relaxed by instead using photometric redshifts, adding random noise, and repeating the upcoming analysis on many realizations of the photometric redshifts.

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Generating Background Galaxies}
With the (0,0,0,0) catalog of MS foreground galaxies chosen, a set of ?? background galaxies over a field of ?? arcmin$^2$ was generated, with density 10 per arcmin$^2$. The intrinsic orientation of each galaxy was sampled from a uniform distribution as, without lensing, there should be no preferred orientation due to the assumption of an isotropic universe. The magnitude of the galaxy ellipticities was sampled from a normal distribution with a standard deviation of 0.2, but any ellipticities with magnitude greater than one were re-sampled. Random ellipticity noise was sampled from a normal distribution with a standard deviation of 0.1 and added to the intrinsic ellipticities. Finally, each ellipticity was multiplied by $M=0.9$ to account for a multiplicative bias of 10\%.

@@

One of the long-term goals of \texttt{Pangloss} is to be able to use all available sky survey data (sky positions, stellar masses, magnitude, photometric redshift, ellipticities, etc.) to make a hierarchial inference of the Universe's cosmological and galaxy population model hyper-parameters such as the density paramters $\Omega_m$ and $\Omega_\Lambda$, the Hubble constant $H_0$, and stellar mass-halo mass relation \cite{smhr}. This is a large undertaking and will require considerable testing and calibration on large N-body simulations such as the Millennium Simulation. As a first step towards this goal, I attempted to gague the feasibility of implementing two methods of parameter inference, maximum likelihood estimation and approximate Bayesian computation, using the current framework. The results of this feasibility test, particularly the CPU time per likelihood estimation, will help guide fugure work on \texttt{Pangloss}. For a detailed description of a probablistic modeling approach to gravitational lensing inference for large sky surveys, see \cite{grav_lens_inference}. Exploratory notes of implementing such an inference in the framework is available on \texttt{Pangloss}'s public github repository\footnote{https://github.com/drphilmarshall/Pangloss/issues/23}.\\

@@

MLE is not the only common method of parameter inference. As discussed in the previous section, the likelihood is computationally expensive to calculate and so MLE may not lend itself well to inference when large number of background galaxies are used. Approximate Bayesian computation (ABC) is a class of computational methods that estimate the posterior distribution without calculating the likelihood function. This can be vital for complex models in which the likelihood function does not have an analytic form or is computationally costly to evaluate. While fortunately Equation \eqref{log-likelihood} is a relatively simple likelihood function, the calculation may become too costly for very large mock catalogs of millions or billions of source galaxies. However, ABC does come at a cost; as the name implies, there are various assumptions and approximations made in the process that can potentially lead to an innaccurate posterior prediction. A general overview of the ABC process is shown in Figure \ref{fig:abc} taken from (first author name et al.) \cite{abc}, which offers an excellent introduction to the topic. I summarize a common ABC implementation and its application to \texttt{Pangloss} below.

@@

ABC using a simple distance metric like Euclidean distance or $\chi^2$ is highly inefficient for data with large dimensionality as the probability of the `distance' being small quickly vanishes. Instead, the similarity of observed and simulated data is often characterized by summary statistics such as mean and variance for simple data or correlation functions for more ... data. Thus an ABC rejection algorithm reduces to predicting the summary statistic for the simulated data, comparing to the corresponding summary statistic of the observed data, and basing rejection on a chosen threshold of difference in summary statistics.

A possible implemenetation of the above for \texttt{Pangloss} would be as follows: Choose the the ray-traced lensed ellipticities as the `observed' data and compute its $\varepsilon$-$\varepsilon$ correlation function to be used as the desired summary statistic. Let the distance metric to be the NRMSE between the predicted and ray-traced $\varepsilon$-$\varepsilon$ correlation function. Compute the best-case $\varepsilon$-$\varepsilon$ correlation function using all halo data as shown in Section \ref{pangloss} and make the threshold value $\delta$ be some multiple of the best-case NRMSE (e.g. ${\delta=2\cdot\text{RMSE}_{\text{best-case}}}$. Create mock data by repeating the methodology in \ref{pangloss} except use the stellar mass-halo mass relation, dependent on paramters we sample from a prior, to predict halo masses based upon each galaxy's stellar mass. Compute the simulated $\varepsilon$-$\varepsilon$ correlation function and determine whether to accept or reject the data by determining if the NRMSE is below the threshold $\delta$. The posterior will be estimated after a large number of iterations through the rejection algorithm.

@@

The results of Section \ref{feasibility} demonstrated that the likelihood calculation does not present a computational issue in comparrison to the lensing calculation, so MLE is a feasible approach for a hierarchial inference of the stellar mass-halo mass relation. A single ABC rejection algorithm in fact performed \textit{worse} than a single likelihood calculation, so clearly ABC is not suitable for this work. However as discussed above, there is still much more work to the \texttt{Pangloss} framework that needs to be done, both in modeling features and computational infastructure, before an inference using a large number of background sources comparable to modern sky surves can be achieved.

@@

OLD: In this thesis, a simple halo model was used to reconstruct lensing mass along lines of sight in the Millennium Simulation to make predictions of how foreground mass weakly lensed the ellipticities of ?? generated background sources across a field of ?? arcmin$^2$. The lensed ellipticities were characterized globally using the ellipticity-ellipticity and galaxy-mass correlation functions and then compared with the same statistics from ray-traced data. There was a systematic-underprediction of shear power on all scales in the ellipticity-ellipticity correlation function as well as an absence of any predicted correlation on scales larger than 1 arcminute, suggesting missing features in the \texttt{Pangloss} framework. The galaxy-mass correlation function predicted by the halo model was in agreement with the ray-traced correlation function on scales from 0.2 to 0.7 arcminutes, but significantly overpredicted the shear power on smaller scales due to strong lensing effects. Now that a proof of concept has been demonstrated, there is work being done to scale up the \texttt{Pangloss} framework to make lensing predictions over 100 deg$^2$ using far more computational resources. While the results are encouraging for such a simple model, many strong and unrealistic assumptions were made in the methodology that limit its use in observational surveys. In the future, a similar analysis should be made after relaxing some of the assumptions to see how different the predictions by the halo model are from this best-case scenario.

>>>>

Predicting accurate dark matter mass maps from weak lensing requires understanding the correlation between galaxies and dark matter structure. Numerous N-body simulations have predicted dark matter structure to have evolved from overdense regions in the early universe into `halos' of virialized matter that eventually collapsed from self-gravity (cite), where galaxies could then form in the large gravitational potentials of the halos. However, there is still extensive reserach being done on the correct density profile of the halos (isothermal spheres \cite{isothermal_spheres}, NFW \cite{nfw}, BMO \cite{bmo}, Einasto \cite{einasto}), identifying the stellar mass - halo mass relation \cite{smhr}, abundance matching \cite{reddick}, halo substructure \cite{substructure}, and the lensing by line-of-sight halos \cite{collet} and by dark matter structures not associated with galaxies \cite{mccully}. If a model for the distribution of dark matter in a region of foreground mass can accurately predict the statistical signal of the weak lensing of background sources, then the model can be used on galaxies in existing sky survey data to extrapolate the amount of dark matter in the region and construct large-scale maps of the dark matter in the universe.

Guided by these results, in this thesis I build upon the work of Collet et al. in \cite{collet} to apply a simple dark matter halo model to reconstruct lensing mass along lines of sight in the Millennium Simulation in order to predict the weakly lensed ellipticities of generated background sources. I then compare my predicted lensed ellipticities to the `true' lensing calculated by Hilbert et al. in \cite{ray_tracing} by ray-tracing light through the Millennium Simulation. By doing so, I attempt to answer the following questions: Under the best possible circumstances of having noiseless redshifts and halo masses, how accurately can I reconstruct line-of-sight dark matter mass structures using a halo model? What fraction of dark matter mass is in halos? How feasible is an inference of the Millennium Simulation's cosmological hyper-parameters, such as the stellar mass - halo mass relation, using all available halo data? How long does such an inference take?

%Guided by these results, in this thesis I build upon the work of Collet et al. in \cite{collet} to apply a simple dark matter halo model to reconstruct lensing mass along lines of sight in the Millennium Simulation using the \texttt{Pangloss} framework in order to predict the weakly lensed ellipticities of generated background sources.

A brief introduction to the halo model, the effects of strong and weak gravitational lensing, representing galaxy ellipticities, and cosmological correlation functions is discussed in Section \ref{background}. The implementation of the \texttt{Pangloss} framework to make line of sight lensing predictions in the Millennium Simulation is described in Section \ref{pangloss}, and the results of the framework on ?? galaxies on a field size of ?? arcmin$^2$, along with a comparison of the predicted lensed ellipticities to ray-traced ellipticities, is given in Section \ref{check}. Section \ref{feasibility} explores the feasibility of using the presented \texttt{Pangloss} framework to make parameter inferences using all available halo data either by likelihood estimation or approximate Bayesian computation. Section \ref{discussion} discusses limitations of the used framework as well as potential physical and computational improvements that can be made for upcoming work before concluding remarks in Section \ref{conclusion}.

@@

Cosmological simulations like the Millennium Simulation (MS) demonstrate that the underlying matter distribution of our universe is likely breathtakingly complex; rather than a largely disordered collection of galaxies and interstellar gas, matter appears to condense into clusters and superclusters of galaxies threaded together by large filaments pocketed with under-dense regions called voids, as shown in Figure \ref{fig:ms_structure} (Cite MS). While the exact relation between the distribution of galaxies and dark matter is not known, simulations predict that early galaxies formed in the large gravitational potentials of over-dense regions of dark matter (Citation). This suggests that, while we cannot map dark matter directly, galaxies should trace out at least some of the underlying dark matter structures (Citation). A comparrison of the predicted galaxy and dark matter distribution from a small region of the MS is shown in Figure \ref{ms_compare}.

@@

The simplest way to model the relationship between galaxies and dark matter is by enveloping each galaxy in a spherically symmetric dark matter `halo' of a certain mass $M_h$. These halos extend far beyond the edge of the visible galaxy that they enclose, with the Milky Way's own halo estimated to have a radius of (Check and cite). While the density profile of the halos may be complex, numerous simulations have shown that it can be well approximated locally by the Navarro-Frenk-White (NFW) profile which has the form
\begin{equation}\label{nfw_profile}
\rho_{NFW}(r)=\frac{\rho_0}{\frac{r}{R_s}\left(1+\frac{r}{R_s}\right)^2}
\end{equation}

\noindent where the constant $\rho_0$ and the scale radius $R_s$ are parameters that vary from halo to halo \cite{nfw}. However, the NFW profile is not entierly physical as the total mass diverges as the desntiy is integrated out to an infinite radius. Instead this work uses a truncated NFW profile called the Baltz-Marshall-Oguri (BMO) profile given by
\begin{equation}\label{bmo_profile}
\rho_{BMO}(r)=\rho_{NFW}\cdot\left(\frac{r_t^2}{r^2+r_t^2}\right)^2
\end{equation}

@@

\noindent where $d_H=c/H_0$ is the Hubble distance, $\Omega_k$ is the curvature parameter ($\Omega_k=0$ for a flat universe), and $d_C(z)$ is the line-of-sight comoving distance given by
\begin{equation}\label{comoving_distance}
d_C(z)=d_H\int_0^z\frac{dz^\prime}{E(z^\prime)}
\end{equation}

@@

As the proper distance between two objects $r_p(z)$ is related to the comoving distance $d_C(z)$ by
\begin{equation}\label{physical2comoving}
r_p(z)=a(z)d_C(z)=\frac{d_H}{1+z}\int_0^z\frac{dz^\prime}{E(z^\prime)},
\end{equation}

@@

\noindent Combining this result with Equation \eqref{temp_solution} leads to the desired proper volume element
\begin{equation}\label{pvolume_element}
dV_p=\frac{d_Hd_M(z)^2}{E(z)(1+z)^3}dzd\Omega.
\end{equation}

@@

A full mathematical treatment of the gravitational lensing of galaxies due to the gravitational fields of massive objects requires general relativity (see \cite{modern_cosmology} for details). However, the important results can be summarized as follows. Foreground mass distorts the image of a background galaxy in two distinct ways: The image is magnified and sheared tangentially about the foreground mass making it more elliptical. The magnification of the image is determined by the convergence $\kappa$ defined as
\begin{equation}\label{convergence}
\kappa(\vec{\theta})=\frac{\sum(\vec{r},z_l)}{\sum_{cr}(z_l,z_s)}
\end{equation}

@@

The quantity usually of interest in lensing calculations is the reduced shear, defined as
\begin{equation}\label{reduced_shear}
g=\frac{\gamma}{1-\kappa}.
\end{equation}

@@

 The analytical form of the convergence and shear contribution from a BMO-profile is quite messy, but is derived and explained in detail in \ref{bmo}. Calculating the lensing contribution of the smooth-component correction is much simpler; the projected surface mass density at at given redshift is found by integrating $\rho_s(z)$ over a sufficiently small $dz$ interval, and then the mass sheet's convergence is found using Equation \eqref{kappa}. There is no shear contribution as a uniform sheet of mass causes only magnification due to the mass sheet degeneracy (CITE!). However, Equation \eqref{lensing} shows that the lensed ellipticities depend on the \textit{reduced} shear, which does depend on the additional convergence terms resulting from the smooth-component correction.

@@

To calculate the lensed ellipticities in the previous section, galaxies must first be translated from pixel intensities into elliptical representations of the galaxies. This is not always possible as galaxies are diverse in type and shape and not all are well approximated by ellipses. However for this thesis, all galaxy ellipticities are simulated so this issue is avoided. For a more thorough analysis of measuring image ellipticities, see \cite{schneider}.

@@

\noindent and $r\leq1$ is the ratio of the semi-minor and semi-major axes of the ellipse. This compact notation combines the eccentricity and orientation of the ellipse into a single object. A plot from \cite{schneider} showing the shape of elliptical galaxies for various values of $\varepsilon_1$ and $\varepsilon_2$ is shown in Figure \ref{ellipses}.

@@

If all galaxies were spherical then the measurement of their ellipticities would give an exact description of the lensing done by foreground mass structures. Unfortunately, the intrinsic ellipticities of background galaxies contribute significant noise as we do not know how any individual galaxy ellipticity has been altered by lensing. However as two sources lensed by the same foreground mass will have their ellipticities distorted in a similar way, their lensed ellipticities should be correlated regardless of intrinsic orientation. The average correlation of a sufficient number of galaxy pairs should overcome the noise and lead to a detectable signal. Importantly, this implies that \textit{the correlation of galaxy ellipticities is a probe of the underlying dark matter mass distribution} (Citation).

@@

As galaxies will only be lensed by the same foreground mass at relatively small separation distances, the $\varepsilon$-$\varepsilon$ correlation function decreases sharply as the separation increases. If the distribution of intrinsic galaxy ellipticities is indeed random, then their $\varepsilon$-$\varepsilon$ correlation function should vanish across all separation scales.

A similar statistical measure often used in weak lensing measurements is the galaxy-mass correlation function which, in this setting, measures the cross-correlation of galaxy ellipticities and halo locations. Simply put, the function measures the correlation of lensed ellipticities around foreground halos as a function of separation distance and only has one component relevant for this work (see \cite{ng_corr} for details).

%  and is defined as
% \begin{equation}\label{ng_corr_def}
% \xi=
% \end{equation}

% \noindent where...\textcolor{red}{Can't seem to find definition??}

@@

To infer foreground dark matter mass structures using weak gravitational lensing, first a model of the relationship between foreground galaxies and dark matter must be established and robustly tested to see if, statistically, it makes the same lensing predictions of background sources as the true underlying dark matter structure. To do this, I built upon the publicly available \texttt{Pangloss}\footnote{\label{note1}https://github.com/drphilmarshall/Pangloss} framework used in Collett et al. (\cite{collett_marshall}) to reconstruct all the
mass along the lines of sight of generated background galaxies in the MS using a halo model with smooth-component correction. The lensing contribution of each halo is calculated, and the total lensing of the background galaxy is the sum of each halo contribution. The implementation of this process is detailed in the following sections.

@@

Testing the first assumption is the main goal of this thesis. The second assumption is reasonable as it is estimated that dark matter halos are on average one to two orders of magnitude more massive than their host galaxies \cite{smhr}. Clearly the third assumption will never be true for any observational data. However, this allows for a best-case estimate of how well the \texttt{Pangloss} framework \textit{could} predict weak lensing effects given all possible information. This assumption can be relaxed by sampling a dark matter halo mass from an assumed stellar mass - halo mass relation. The fourth assumption is also unrealistic as most galaxies in sky surveys only have a less-reliable photometric redshift due to time constraints, but this again allows for a best-case estimate. This assumption could be relaxed by instead using photometric redshifts, adding random noise, and repeating the upcoming analysis on many realizations of the photometric redshifts.

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Generating Background Galaxies}
With the (0,0,0,0) catalog of MS foreground galaxies chosen, a set of ?? background galaxies over a field of ?? arcmin$^2$ was generated, with density 10 galaxies per arcmin$^2$. The intrinsic orientation of each galaxy was sampled from a uniform distribution as, without lensing, there should be no preferred orientation due to the assumption of an isotropic universe. The magnitude of the galaxy ellipticities was sampled from a normal distribution with a standard deviation of 0.2, but any ellipticities with magnitude greater than one were re-sampled. Random ellipticity noise was sampled from a normal distribution with a standard deviation of 0.1 and added to the intrinsic ellipticities. Finally, each ellipticity was multiplied by $M=0.9$ to account for a multiplicative bias of 10\%.

@@

One of the long-term goals of \texttt{Pangloss} is to be able to use all available sky survey data (sky positions, stellar masses, magnitude, photometric redshift, ellipticities, etc.) to make a hierarchial inference of the Universe's cosmological and galaxy population model hyper-parameters such as the density paramters $\Omega_m$ and $\Omega_\Lambda$, the Hubble constant $H_0$, and stellar mass - halo mass relation \cite{smhr}. This is a large undertaking and will require considerable testing and calibration on large N-body simulations such as the Millennium Simulation. As a first step towards this goal, I attempted to gague the feasibility of implementing two methods of parameter inference, maximum likelihood estimation and approximate Bayesian computation, using the current framework. The results of this feasibility test, particularly the CPU time per likelihood estimation, will help guide fugure work on \texttt{Pangloss}. For a detailed description of a probablistic modeling approach to gravitational lensing inference for large sky surveys, see \cite{grav_lens_inference}. Exploratory notes of implementing such an inference in the framework is available on \texttt{Pangloss}'s public github repository\footnote{https://github.com/drphilmarshall/Pangloss/issues/23}.\\

@@

MLE is not the only common method of parameter inference. As discussed in the previous section, the likelihood is computationally expensive to calculate and so MLE may not lend itself well to inference when large number of background galaxies are used. Approximate Bayesian computation (ABC) is a class of computational methods that estimate the posterior distribution without calculating the likelihood function. This can be vital for complex models in which the likelihood function does not have an analytic form or is computationally costly to evaluate. While fortunately Equation \eqref{log-likelihood} is a relatively simple likelihood function, the calculation may become too costly for very large mock catalogs of millions or billions of source galaxies. However, ABC does come at a cost; as the name implies, there are various assumptions and approximations made in the process that can potentially lead to an innaccurate posterior prediction. A general overview of the ABC process is shown in Figure \ref{fig:abc} taken from \cite{abc}, which offers an excellent introduction to the topic. I summarize a common ABC implementation and its application to \texttt{Pangloss} below.

@@

ABC using a simple distance metric like Euclidean distance or $\chi^2$ is highly inefficient for data with large dimensionality as the probability of the `distance' being small quickly vanishes. Instead, the similarity of observed and simulated data is often characterized by summary statistics such as mean and variance for simple data or correlation functions for more (advanced) data. Thus an ABC rejection algorithm reduces to predicting the summary statistic for the simulated data, comparing to the corresponding summary statistic of the observed data, and basing rejection on a chosen threshold of difference in summary statistics.

A possible implemenetation of the above for \texttt{Pangloss} would be as follows: Choose the the ray-traced lensed ellipticities as the `observed' data and compute its $\varepsilon$-$\varepsilon$ correlation function to be used as the desired summary statistic. Let the distance metric to be the NRMSE between the predicted and ray-traced $\varepsilon$-$\varepsilon$ correlation function. Compute the best-case $\varepsilon$-$\varepsilon$ correlation function using all halo data as shown in Section \ref{pangloss} and make the threshold value $\delta$ be some multiple of the best-case NRMSE (e.g. ${\delta=2\cdot\text{RMSE}_{\text{best-case}}}$). Create mock data by repeating the methodology in \ref{pangloss} except use the stellar mass - halo mass relation, dependent on paramters we sample from a prior, to predict halo masses based upon each galaxy's stellar mass. Compute the simulated $\varepsilon$-$\varepsilon$ correlation function and determine whether to accept or reject the data by determining if the NRMSE is below the threshold $\delta$. The posterior will be estimated after a large number of iterations through the rejection algorithm.

@@

The results of Section \ref{feasibility} demonstrated that the likelihood calculation does not present a computational issue in comparrison to the lensing calculation, so MLE is a feasible approach for a hierarchial inference of the stellar mass - halo mass relation. A single ABC rejection algorithm in fact performed \textit{worse} than a single likelihood calculation, so clearly ABC is not suitable for this work. However as discussed above, there is still much more work to the \texttt{Pangloss} framework that needs to be done, both in modeling features and computational infastructure, before an inference using a large number of background sources comparable to modern sky surves can be achieved.

@@

%OLD: In this thesis, a simple halo model was used to reconstruct lensing mass along lines of sight in the Millennium Simulation to make predictions of how foreground mass weakly lensed the ellipticities of ?? generated background sources across a field of ?? arcmin$^2$. The lensed ellipticities were characterized globally using the ellipticity-ellipticity and galaxy-mass correlation functions and then compared with the same statistics from ray-traced data. There was a systematic-underprediction of shear power on all scales in the ellipticity-ellipticity correlation function as well as an absence of any predicted correlation on scales larger than 1 arcminute, suggesting missing features in the \texttt{Pangloss} framework. The galaxy-mass correlation function predicted by the halo model was in agreement with the ray-traced correlation function on scales from 0.2 to 0.7 arcminutes, but significantly overpredicted the shear power on smaller scales due to strong lensing effects. Now that a proof of concept has been demonstrated, there is work being done to scale up the \texttt{Pangloss} framework to make lensing predictions over 100 deg$^2$ using far more computational resources. While the results are encouraging for such a simple model, many strong and unrealistic assumptions were made in the methodology that limit its use in observational surveys. In the future, a similar analysis should be made after relaxing some of the assumptions to see how different the predictions by the halo model are from this best-case scenario.
